---
title: "Introduction to the net.raster package and its functionalities"
author: "Cynthia Valéria Oliveira and Neander Marcel Heming"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the net.raster package and its functionalities}
  %\usepackage[utf8]\n 
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: sentence  
---

```{r setup, echo = FALSE, include=FALSE}
library(knitr)
knitr::opts_chunk$set(collapse = TRUE,message=FALSE, warning = FALSE,
                      comment = "#>")
```

-   [Introduction](#intro)
-   [Installation](#install)
-   [Data processing](#data)
-   [Analysis](#analysis)
-   [References](#references)

## Introduction {#intro}

[`net.raster`](https://github.com/CynOliveira/net.raster) is an R package to 
calculate spatialized interaction network metrics for bipartite networks, using species presence-absence rasters as input, in addition to the interaction matrix. Thus, it improves the visualization of interaction patterns in space, generating a new layer of information based on species distributions. We can calculate potential bipartite interactions on a grid by combining Species Distribution Models (SDM), or even other species distribution raster data, from each trophic level. `net.raster` also allows simultaneous calculation of network metrics for multiple potential communities in space or spatial scenarios over time. The package can be used for mutualistic or antagonistic interactions.

`net.raster` ecological network calculations are based on three steps. Initially, we need to carry out and/or compilate field studies that record interactions between species; they can be binary (interactions presence-absence) or quantitative (usually interaction frequency). We then use species distributions (such as those estimated through ecological niche models) of each of the interacting species for the whole area of interest. At this step, we can also use other types of presence-absence distribution maps, like those provided by IUCN. Finally, it is necessary to combine the distribution rasters of all species at both trophic levels with the information about their interactions. 

The initial step requires combining all species distribution rasters of each trophic level (lower and higher level - LL and HL) on two raster stacks (multi-layer SpatRaster objects from terra R package, Hijmans 2022). Another required input is the network matrix, which records observed interactions. In the next step the `prep.web()` function verifies if the species names in the raster stacks of both trophic levels match the species names in the interaction matrix. Otherwise, an error message will appear, and the user must review the inputs.

Following the bipartite R package (Dormann et al. 2008, 2009; Dormann 2011) original features, the `net.raster` package allows users to calculate network topography indices in three different ways: a) the entire web (default) using the `networklevel.spat()` function, b) each group level, selecting either the lower or higher level in this same function through the argument “level”, or c) for each species, with the `specieslevel.spat()` function, also choosing both (default) or one level of interest (higher lower) at a time. Users can also choose unweighted metrics whenever their calculation allows, default being the weighted calculation for `networklevel.spat()` and unweighted computation for `specieslevel.spat()`. 

There are currently 15 indices that can be calculated using the function `networklevel.spat()` through the “index” argument: connectance, weighted connectance, web specialisation (H2’), web asymmetry, links per species, linkage density, number of compartments, compartment diversity, cluster coefficient, ISA (or alternatively interaction strength asymmetry or dependence asymmetry), SA (or alternatively specialisation asymmetry), Fisher alpha, interaction evenness, Alatalo interaction evenness, and Shannon diversity. These indices are equivalent to the ones from the `networklevel()` function of the bipartite R package. Also, `networklevel.spat()` provides four metrics for nestedness, including NODF (an acronym for weighted Nested Overlap and Decreasing Fill), weighted nestedness, and weighted NODF. Moreover, there are those metrics invoked through “grouplevel” (higher or lower trophic level): number of species in the respective trophic level, mean number of links, mean number of shared partners, weighted cluster coefficient, degree distribution, togetherness, C score, V ratio, discrepancy, extinction slope, robustness, niche overlap, generality, vulnerability, and fc (or alternatively functional complementarity).

The `specieslevel.spat()` metrics also follow arguments from the `specieslevel()` function of bipartite R package. The current 18 indices are: degree, normalised degrees, species strength, nestedrank as rank in a nested matrix, interaction push pull, PDI for Paired Differences Index, resource range, species specificity (or coefficient of variation of interactions), PSI for pollination service index (or pollinator support index, depending on the trophic level), NS for node specialisation index, betweenness (also returns its weighted counterpart), closeness (also returns its weighted counterpart), Fisher for Fisher's alpha index, partner diversity for Shannon diversity of interactions, effective partners, proportional generality (a quantitative version of normalised degree), proportional similarity (specialisation measured as similarity between use and availability), and d for Blüthgen's d' (Dormann et al. 2008). 

Other two functions provided by `net.raster` are related with two important, structural network-level measures: modularity and nestedness. The `computeModules.spat()` function computes modules of a bipartite weighted graph by applying Newman's modularity measure (Newman and Girvan, 2004), while the `nested.spat()` function provides additional features compared to the `networklevel.spat()` function by giving the results of three other nesting metrics and expanding the possibilities based on different null models and theoretical approaches for this web property. As output, in the case of community-level functions, the raster will have as many layers as metrics were required, while the `specieslevel.spat()` function generates a layer for each species in each indicated metric. 

The `net.raster` R package is the first tool capable of calculating bipartite ecological network metrics by combining binary species distribution rasters and interaction matrices. `net.raster` allows the spatial calculation and visualisation of interaction diversity metrics with efficient memory usage. It is based primarily on extrapolation (spatial or temporal) of actual records of interactions, thus using co-occurrence as a proxy for an interaction between species. Despite some criticism of this approach, co-occurrence analysis may reveal new patterns of spatial association between pairs of known species (Thurman et al. 2019), possibly opening up other questions in the macroecological and biogeographic study of networks. In addition, an accurate projection based on the mapping of true interactions seems impractical at least for now (Poisot et al. 2021). However, `net.raster` will be useful to help accessing geographical differences between actual and projected interactions to refine our knowledge of the causes of macroecological patterns in species interaction networks. Advances in mapping interactions and expanding field studies of interaction data around the world can provide new forms of spatial information directly associated with interactions. Also, `net.raster` is flexible to use raster data other than those derived from species distribution models (SDMs), and we encourage researchers to collaborate with us for the improvement of this free and pioneering tool in the development of the research field of macroecology and biogeography of diversity of interactions. In this vignette, we demonstrate the use of the functions of the `net.raster`.
package in detail.

## Installation {#install}

The development version of `net.raster` can be downloaded from [GitHub](https://github.com/CynOliveira/net.raster) using the following code:
```{r, eval = FALSE}
devtools::install_github("CynOliveira/net.raster")
```

If you have any questions, let us know through the topic ["Issues"](https://github.com/CynOliveira/net.raster/issues).

`net.raster` uses some R packages as dependencies such as `bipartite` (version >= 3.5.0) (Dormann et al. 2008) and `terra` (version \>= 1.6) (Hijmans, 2022).
Once installed, packages can be loaded into R using `library()`:

```{r, warning = FALSE, message = FALSE}
library(bipartite)
library(terra)

```

## Data processing {#data}
In the step of pre processing, we offer support to prepares raster and network data, excluding in the network the species that not interact. The function `prep.web()`
also checks whether the species names of each Spatraster (higher and lower level) match the species names in the cleaned network.
In the processing step, we provide functions to calculate network topography indices using: a) the entire web, b) selecting the lower or upper level of each group, or c) selecting each species, choosing both levels or one level of interest at a time. Some indices calculated by net.raster are connectance, interaction evenness, web specialisation (H2’) and degree distribution, besides species-level metrics such as  species strength, betweenness and closeness centralities, degrees, and species specialisation (d’). `net.raster` also allows to calculate, in different ways, two important network-level structural measures: nestedness and modularity. 
Note that if a network is very small, with few nodes or links, it may be impossible to calculate the choose metric or calculated result may be unreliable.As the calculation is made with the subnet of each pixel, even in cases like this, it is possible to visualize the spatialized metric on a macroecological scale. 

-   **Dataset**

We demonstrate the applicability of the net.raster by using a dataset of a real plant-frugivore network recorded in the Brazilian Atlantic Forest (Silva et al. 2002). We subsetted the original dataset to 70 bird species and 138 plant species. This subsetted network contains only those bird and plant species whose modelling was possible, due to the number of occurrence points. We provide the quantitative interaction matrix and the species distribution rasters, projected for current climate variables across the Atlantic Forest using MaxEnt algorithm (Phillips et al. 2006, 2017) through ENMWizard R package (Heming et al. 2018). 

```{r}
# load bipartite network and the raster stacks of higher level and lower level species
rasth <- terra::rast(system.file("extdata", "rasth.tif",
                            package="net.raster"))
rastl <- terra::rast(system.file("extdata", "rastl.tif",
                            package="net.raster"))
bipnet <- read.csv(system.file("extdata", "bipnet.csv",
                                  package="net.raster"), row.names=1, sep= ";" )
```

-   **Function** `prep.web`

This function prepares raster and network data, excluding in the network the species that not interact. Also checks whether the species names of each Spatraster (higher and lower level) match the species names in the cleaned network.

```{r}
# applying the function to check species names on all data
resu  <- prep.web (rasth, rastl, bipnet)
resu
```

## Analysis {#analysis}

-   **Function** `nested.spat`
This function calculates any of several nestedness metrics available for bipartite networks, which can be weighted or not, for raster data. For a time-efficient processing of spatial data, users may choose one of the nine available nestedness indices ("discrepancy", "discrepancy2", "binmatnest", "NODF", "NODF2", "C score", "checker","weighted NODF", "wine"). See more about them at \link[bipartite]{nested}

```{r}
# applying the function to compute weighted NODF (default)
wNODF <- nested.spat (rasth, rastl, bipnet)
plot(wNODF)
# applying the function to compute C score index
Cscore <- nested.spat (rasth, rastl, bipnet, method = "C score")
plot(Cscore)
```

-   **Function** `computeModules.spat`
This function calculates modules by applying Newman's modularity measure to a spatial weighted bipartite network. Users must choose between the algorithms of Stephen Beckett (2016) or Dormann & Strauss (2016) (method="DormannStrauss"). The default is the Beckett algorithm, which is faster and generally better. View more about it at  \link[bipartite]{computeModules}

```{r}
# computing Newman's modularity with faster method, Beckett algorithm (default)
compMod <- computeModules.spat (rasth, rastl, bipnet)
plot(compMod)
# calculating Newman's modularity with alternative method
compModDS <- computeModules.spat (rasth, rastl, bipnet, method="DormannStrauss")
plot(compModDS)
```

-   **Function** `specieslevel.spat`
This function calculates various simple numbers and indices related to network nodes (species) for raster data. You can see the spatial variation of important metrics such as species strength and centralities (closeness - default metric, betweenness, and degree). View all available indexes in Details and note that users may select one index at a time for time-efficient spatial calculation. For levels, "both" is the default, but users can also choose one level of interest at a time (higher or lower), as well as choose weighted metrics whenever their calculation allows, with the default being unweighted calculation.
Note that if a network is very small, with few nodes or links, it may be impossible to calculate the chosen metric or the calculated result may be unreliable. For example, some metrics, such as closeness and betweenness centrality, may present an error message for some pixels, but be computed over the spatial extent in all pixels where this is possible. As the calculation is made with the subnetwork of each pixel, even in these cases it is possible to visualize the spatialized metric on a macroecological scale. 
Users may select one index at a time for time-efficient spatial calculation, there will be an error message otherwise. Also note that some indices may have a high processing time for the entire raster, depending on their algorithms, the size of the rasters and the network, and the processing capacity and available RAM of the machine used.
The current available indices for species level metrics are listed bellow and users can find more information about them at  \link[bipartite]{specieslevel}
• ‘degree’,
• 'normalised degree',
• ‘species strength’,
• ‘nestedrank’,
• ‘interaction push pull’ - interaction push/pull
• ‘PDI’ - Paired Differences Index,
• ‘resource range’,
• ‘species specificity’,
• ‘PSI’ - pollination service index ,
• ‘betweenness’ - betweenness centrality and its weighted counterpart,
• ‘closeness’ - closeness centrality and its weighted counterpart,
• ‘Fisher alpha’ - Fisher’s alpha index,
• ‘partner diversity’ - Shannon diversity of interactions
• ‘effective partners’,
• ‘proportional generality’,
• ‘proportional similarity’,
• ‘d’ - Blüthgen’s d’

```{r}
# applying the function to compute closeness centrality (default metric) for lower level
lowclos <- specieslevel.spat (rasth, rastl, bipnet, level="lower")
plot(lowclos)
plot(lowclos$abuta_selloana) 
# you can choose one species at a time for better visualization
# calculating closeness, now for both levels (default)
allclos <- specieslevel.spat (rasth, rastl, bipnet)
plot(allclos$attila_rufus) # you can choose one species at a time
plot(allclos$abuta_selloana) #of each level
# computing normalised degree for higher level
hignd <- specieslevel.spat (rasth, rastl, bipnet, index="normalised degree",
level="higher")
plot(hignd$saltator_similis)# you can choose one species at a time
```

-   **Function** `networklevel.spat` 
This function calculates various indices and values for bipartite networks on raster data. You can see the spatial variation of important metrics such as connectance, web asymmetry, network specialization (H2) and four options to compute nestedness. 
Users may select one index at a time for time-efficient spatial calculation, being the default the "connectance". Also note that some indices may have a high processing time for the entire raster, depending on their algorithms and the processing capacity and RAM of the machine used. The current available indices for network level metrics are listed bellow and users can find more information about them at \link[bipartite]{networklevel}.

There are metrics computed for the entire network:
• ‘connectance’,
• ‘web asymmetry’,
• ‘links per species’,
• ‘number of compartments’,
• ‘compartment diversity’,
• ‘cluster coefficient’,
• ‘nestedness’
• ‘NODF’,
• ‘weighted nestedness’
• ‘weighted NODF’,
• ‘ISA’ (or alternatively ‘interaction strength asymmetry’ or ‘dependence
asymmetry’),
• ‘SA’ (or alternatively ‘specialisation asymmetry’),
• ‘linkage density’,
• ‘weighted connectance’,
• ‘Fisher alpha’,
• ‘interaction evenness’,
• ‘Alatalo interaction evenness’,
• ‘Shannon diversity’,
• ‘H2’ (network specialization);

and/or those metrics invoked through "grouplevel":
• ‘number of species’ in the respective trophic level,
• ‘mean number of links’,
• ‘mean number of shared partners’,
• ‘weighted cluster coefficient’,
• ‘degree distribution’,
• ‘togetherness’,
• ‘C score’,
• ‘V ratio’,
• ‘discrepancy’,
• ‘extinction slope’,
• ‘robustness’,
• ‘niche overlap’,
• ‘generality’,
• ‘vulnerability’,
• ‘fc’ (or alternatively ‘functional complementarity’).

```{r}
# applying the function to compute connectance (default) for both levels (default), that is, the entire network
conn <- networklevel.spat (rasth, rastl, bipnet)
plot(conn)
# calculating  H2 for he entire network
webh2 <- networklevel.spat (rasth, rastl, bipnet, index="H2")
plot(webh2)
# computing degree distribution for one group, the lower level
lowdd <- networklevel.spat (rasth, rastl, bipnet, index="degree distribution",
level="lower")
plot(lowdd)
```

## References {#references}

Dormann, C.F., Fruend, J., Bluethgen, N. & Gruber B. 2009. Indices, graphs and null models: analyzing bipartite ecological networks. The Open Ecology Journal, 2, 7-24.

Dormann, C.F., Gruber B. & Fruend, J. (2008). Introducing the bipartite Package: Analysing Ecological Networks. R news Vol 8/2, 8 - 11.

Dormann, C.F. (2011). How to be a specialist? Quantifying specialisation in pollination networks. Network Biology 1, 1 - 20.

Heming, N. M.; Dambros, C.; Gutiérrez, E. E. (2018). ENMwizard: AIC model averaging and other advanced techniques in Ecological Niche Modeling made easy. <https://github.com/HemingNM/ENMwizard>

Hjimans, R. J. (2022). Terra, Spatial Data Analysis (1.6.7).       <https://CRAN.R-project.org/package=terra>

Newman,M. E. & Girvan, M.( 2004).  Finding and evaluating community structure in networks. Physical review E, 69(2), p. 026113. <https://doi.org/10.1103/PhysRevE.69.026113>

Phillips SJ, Anderson RP, Dudík M, et al (2017) Opening the black box: an open-source release of Maxent. Ecography, 40, 887–893. <https://doi.org/10.1111/ecog.03049>

Phillips SJ, Anderson RP, Schapire RE (2006) Maximum entropy modeling of species geographic distributions. Ecological Modelling, 190, 231–259. <https://doi.org/10.1016/j.ecolmodel.2005.03.026>

Poisot, T., Bergeron, G., Cazelles, K., Dallas, T., Gravel, D., MacDonald, A., Mercier, B., Violet, C., Vissault, S. (2021). Global knowledge gaps in species interaction networks data. Journal Of Biogeography, 48(7), 1552-1563. <http://dx.doi.org/10.1111/jbi.14127>

Silva, W. R. et al. (2002). Patterns of fruit – frugivore bird communities of South-Eastern Brazil: implications for conservation. Seed Dispersal and Frugivory: Ecology, Evolution and Conservation, 423–435. <https://doi.org/10.1079/9780851995250.0423>

Thurman, L., Barner, A.K., Garcia, T.S., Chestnut, T. (2019). Testing the link between species interactions and species co-occurrence in a trophic network. Ecography, 42(10), 1658-1670. <https://doi.org/10.1111/ecog.04360>

