---
title: "b-vignette"
author: Cynthia Oliveira
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{b_function-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
# Introduction
The goal of *renm* is rasterizing network metrics for ecological and macroecological purposes, initially focusing on bipartite networks of mutualistic interactions.
***
## Functions
### Interaction strength (b)
This function calculates interaction strength for each specie in the network.
***
## Creating random data to test functions of the package
### Calculating network metrics
net1 <- matrix(c(0, 1, 1, 1, 0, 1, 1, 1, 0, 0, 0, 1, 1, 1, 0, 1), nrow = 4, ncol = 4, dimnames = list(c("B1", "B2", "B3", "B4"), c("P1", "P2", "P3", "P4")))
b(net1)
AS(net1)
***
### Spatializing interactions occurrences [in construction]
library(terra)
From a binary matrix with columns being "longitude", "latitude" and then all pairs of interacting species or an inverse edgelist of interactions aggregated to the coordinates, 
datarm <- as.data.frame(yourmatrix, xy=TRUE)
spac_inter <- rast(datarm, type = "xyz")
names(sitesr) <- colnames(datarm[,-c(1,2)])
writeRaster(sitesr, "sites_inter.tif", overwrite = TRUE)
```{r}


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(renm)
```
